SunPy - Python for Solar Physics

### 2014-01-29 1-Introduction.tex
closes issue #75

Science is driven by the analysis of data. Modern advances in sensor
technology, combined with the availability of inexpensive storage has led to
rapid increases in amount of data available to scientists in almost
every discipline.  Solar physics is no exception to this trend. For example, NASA's
Solar Dynamics Observatory (SDO) satellite, launched in February 2010,
produces over 1 TB of catalogued and stored data per day [REF]. Managing and
analysing this mountain of data
requires increasingly sophisticated software tools.  Key qualities
that these tools should have include robustness, ease of use, a
transparent development history, ease of modification, and conformance
to standards.  Software with these qualities provides a strong
foundation can be responsive to the needs of the community as data
volumes grow and science questions evolve.
%schriste - the opening paragraph still needs work
%ji - I had a go.

>>>>

Science is driven by the analysis of data. Modern advances in sensor
technology, combined with the availability of inexpensive storage has led to
rapid increases in amount of data available to scientists in almost
every discipline.  Solar physics is no exception to this trend. For example, NASA's
AIA instrument on the Solar Dynamics Observatory (SDO) satellite, launched in February 2010,
produces over 1 TB of catalogued and stored data per day \citep{aia}. Managing and
analysing this mountain of data
requires increasingly sophisticated software tools.  Key qualities
that these tools should have include robustness, ease of use, a
transparent development history, ease of modification, and conformance
to standards.  Software with these qualities provides a strong
foundation can be responsive to the needs of the community as data
volumes grow and science questions evolve.
%schriste - the opening paragraph still needs work
%ji - I had a go.

### 2014-01-29 3-4-Helioviewer.tex
closes issue #75

SunPy provides the ability to download images hosted by the
\href{http://helioviewer.org}{Helioviewer Project}.  
The aim of the Helioviewer Project is to enable
the exploration of solar and heliospheric data from multiple data
sources (such as instrumentation and feature/event catalogues) via
easy-to-use visual interfaces.  The primary data provided by the
Helioviewer Project are images in the JPEG2000 format (for further
details on the Helioviewer Project see \cite{muller2009}). The
JPEG2000 files are typically highly compressed compared to the source
FITS files they are generated from, and so can be a used as a shortcut
to visualise large amounts of data from multiple sources.  SunPy is
also used in Helioviewer production servers to manage the download and
ingestion of JPEG2000 files from remote servers.

>>>>

SunPy provides the ability to download images hosted by the
\href{http://helioviewer.org}{Helioviewer Project}.  
The aim of the Helioviewer Project is to enable
the exploration of solar and heliospheric data from multiple data
sources (such as instrumentation and feature/event catalogues) via
easy-to-use visual interfaces.  The primary data provided by the
Helioviewer Project are images in the JPEG2000 format (for further
details on the Helioviewer Project see \citep{muller2009}). The
JPEG2000 files are typically highly compressed compared to the source
FITS files they are generated from, and so can be a used as a shortcut
to visualise large amounts of data from multiple sources.  SunPy is
also used in Helioviewer production servers to manage the download and
ingestion of JPEG2000 files from remote servers.

### 2014-01-29 5-Dev.tex
closes issue #75

The continued success of an open-source project depends on many factors;
three of the most important are (1) utility and quality of the code, (2) documentation, and (3) an
active community \cite{bangerth2013}. Several tools (some specific to Python) are used by
SunPy to make achieving these goals more accessible. To maintain high-quality code, a 
transparent and collaborative development workflow made possible by GitHub is used (described
further below).
Additionally, SunPy makes use of `continuous integration' provided by
Travis CI (\url{http://travis-ci.org}), a process by which the addition of any new code 
automatically triggers a comprehensive review of written unit tests. If any single test
fails then the community is alerted before the code is accepted. The unit test coverage is
also monitored by the \href{http://coveralls.io}{Coveralls} service.

>>>>

The continued success of an open-source project depends on many factors;
three of the most important are (1) utility and quality of the code, (2) documentation, and (3) an
active community \citep{bangerth2013}. Several tools (some specific to Python) are used by
SunPy to make achieving these goals more accessible. To maintain high-quality code, a 
transparent and collaborative development workflow made possible by GitHub is used (described
further below).
Additionally, SunPy makes use of `continuous integration' provided by
Travis CI (\url{http://travis-ci.org}), a process by which the addition of any new code 
automatically triggers a comprehensive review of written unit tests. If any single test
fails then the community is alerted before the code is accepted. The unit test coverage is
also monitored by the \href{http://coveralls.io}{Coveralls} service.

### 2014-01-29 solarsoft.tex
repaired i.e. and e.g.
closes issue #75

\section{Comparison to SolarSoft}\label{sec:ssw}
This section compares the SunPy library that is described in Sections 
\ref{sec:DataTypes}--\ref{sec:util} with the current de-facto standard solar 
physics data analysis library: SolarSoft (SSW) \citep{freeland1998}.
%ARI - is is SolarSoft or SolarSoftWare?! The latter includes the W :-)
SolarSoft is a set of integrated software libraries, databases, and system 
utilities which provide a common programming and data-analysis environment for 
solar physics. SSW is modular in nature, which allows users to add required 
packages which contain specialised software for different instrument or tools.

@@

SSW relies upon IDL (Interactive Data 
Language), a commercial and closed-source interactive data analysis environment 
sold by Exelis, a global aerospace, defence, information and services company. 
The SSW `gen' package provides the base analysis capabilities such as time-series
analysis, time conversions, spectral fitting, image display, and file 
I/O. It is composed of 89\% IDL code while the remainder is mostly Perl (5.4\%) 
and csh scripts (3.8\%). 
According to data generated using David A. Wheeler's `SLOCCount'\footnote{\url{http://www.dwheeler.com/sloccount/}}, SSW/gen is composed of 
289,724 lines of code with a total approximate value of \$10.4M (based on the 
Basic COCOMO model \citep{_cocomo_2014}). 
%dps - This data was calculated using...} If we can add them to the repository 
%will improve the quality of the paper and we can always repeat the calculation in the future.
%schriste - i don't know what you mean by this 
%Stuart: comment was Davids
Individual missions and instruments provide additional optional packages to 
support and process their data. For example, the SDO package \citep{sdo} is 
100\% IDL code and is 11k lines of code (\$360k) and the SOHO package 
\citep{soho} is 100k lines of code (\$3M) and is 99.9\% IDL code. SSW is hosted 
and distributed by the Lockheed Martin Solar and Astrophysics Laboratory.
%The RHESSI package is 99.97\% IDL code and is 145k total lines of code (\$5M)
While SSW itself is open-source and freely available, the IDL core is not. The cost 
of IDL is prohibitive for many undergraduates as well as institutions 
with limited means.
Even though SSW's source is open, the development of the source is primarily closed,
with no open issue tracker or established mechanism for community contributions.

One of SunPy's key aims is to provide a free and modern alternative to the 
SolarSoft (gen) library. This is made possible by the recent rise of the 
scientific Python community. Python, as described in Section \ref{sec:Intro}, 
is a general purpose high-level programming language which is free to use and 
cross-platform and powerful programming language.
It therefore has a very large user base which extends far outside of the 
solar physics or astronomy communities. It is very extensible with C, C++, 
FORTRAN or even IDL. Many books and on-line documentation sources are available. For 
example, at the time writing a search 
%
%dps - \footnote{We need to be careful with this! IDL means more things that 
%our 
%beloved Interactive Data Language. If we search for that and programming 
%then we get just 7 books! - from which 5 are just the same. I don't know how we 
%could make a more accurate search.
%If we script it it would be amazing!  Also, I would just provide the order
%of magnitude, and not a exact number, i.e. 190 and 1800 results}
% 
% schriste - I think we defined IDL properly. I was being generous with the 
%book search for IDL
%Stuart: again not not mine, I think we should drop the book thing, in 
%comparison to the other metrics used it's a little weak.
on Amazon for books on IDL programming generated $189$ results while a search 
for Python yielded $1\,788$ results. Python has a strong emphasis on 
readability, a concept which is well-aligned with the scientific endeavour.
%In fairness, IDL is also quite readable! Could move this statement to the intro instead.
Additionally, Python is increasingly being taught in universities, which means 
that research projects can begin quickly without the need for students to learn 
a new (and relatively niche) language. Finally, knowledge of Python is a skill 
which is useful and well compensated outside of the scientific community 
(average salary of \$110k in the Washington, DC, area according to indeed.com at 
% RJH: Is this for a python developer?  Again, this is useless info, imho.
the time of writing). Python continues to see increased use in the astronomy 
community \citep{greenfield2011} which has similar goals and requirements as 
the solar physics community. Finally, Python integrates well with many 
technologies such as web servers \citep{dolgert2008}, SQL databases, 
interactive and shareable notebook-based computing \citep{perez2007}. For all 
of these reasons, Python is an ideal platform for a modern data analysis environment 
for solar physics.

>>>>

\section{Comparison to SolarSoft}\label{sec:ssw}
This section compares the SunPy library that is described in Sections 
\ref{sec:DataTypes}--\ref{sec:util} with the current de-facto standard solar 
physics data analysis library: SolarSoft (SSW) \citep{ssw}.
%ARI - is is SolarSoft or SolarSoftWare?! The latter includes the W :-)
SolarSoft is a set of integrated software libraries, databases, and system 
utilities which provide a common programming and data-analysis environment for 
solar physics. SSW is modular in nature, which allows users to add required 
packages which contain specialised software for different instrument or tools.

@@

SSW relies upon IDL (Interactive Data 
Language), a commercial and closed-source interactive data analysis environment 
sold by Exelis, a global aerospace, defence, information and services company. 
The SSW `gen' package provides the base analysis capabilities such as time-series
analysis, time conversions, spectral fitting, image display, and file 
I/O. It is composed of 89\% IDL code while the remainder is mostly Perl (5.4\%) 
and csh scripts (3.8\%). 
According to data generated using David A. Wheeler's `SLOCCount'\footnote{\url{http://www.dwheeler.com/sloccount/}}, SSW/gen is composed of 
289,724 lines of code with a total approximate value of \$10.4M (based on the 
Basic COCOMO model \citep{cocomo}). 
%dps - This data was calculated using...} If we can add them to the repository 
%will improve the quality of the paper and we can always repeat the calculation in the future.
%schriste - i don't know what you mean by this 
%Stuart: comment was Davids
Individual missions and instruments provide additional optional packages to 
support and process their data. For example, the \texttt{sdo} package \citep{sdo} is 
100\% IDL code and is 11k lines of code (\$360k) and the \texttt{soho} package 
\citep{soho} is 100k lines of code (\$3M) and is 99.9\% IDL code. SSW is hosted 
and distributed by the Lockheed Martin Solar and Astrophysics Laboratory.
%The RHESSI package is 99.97\% IDL code and is 145k total lines of code (\$5M)
While SSW itself is open-source and freely available, the IDL core is not. The cost 
of IDL is prohibitive for many undergraduates as well as institutions 
with limited means.
Even though SSW's source is open, the development of the source is primarily closed,
with no open issue tracker or established mechanism for community contributions.

One of SunPy's key aims is to provide a free and modern alternative to the 
SolarSoft (gen) library. This is made possible by the recent rise of the 
scientific Python community. Python, as described in Section \ref{sec:Intro}, 
is a general purpose high-level programming language which is free to use and 
cross-platform and powerful programming language.
It therefore has a very large user base which extends far outside of the 
solar physics or astronomy communities. It is very extensible with C, C++, 
FORTRAN or even IDL. Many books and on-line documentation sources are available. For 
example, at the time writing a search 
%
%dps - \footnote{We need to be careful with this! IDL means more things that 
%our 
%beloved Interactive Data Language. If we search for that and programming 
%then we get just 7 books! - from which 5 are just the same. I don't know how we 
%could make a more accurate search.
%If we script it it would be amazing!  Also, I would just provide the order
%of magnitude, and not a exact number, i.e., 190 and 1800 results}
% 
% schriste - I think we defined IDL properly. I was being generous with the 
%book search for IDL
%Stuart: again not not mine, I think we should drop the book thing, in 
%comparison to the other metrics used it's a little weak.
on Amazon for books on IDL programming generated $189$ results while a search 
for Python yielded $1\,788$ results. Python has a strong emphasis on 
readability, a concept which is well-aligned with the scientific endeavour.
%In fairness, IDL is also quite readable! Could move this statement to the intro instead.
Additionally, Python is increasingly being taught in universities, which means 
that research projects can begin quickly without the need for students to learn 
a new (and relatively niche) language. Finally, knowledge of Python is a skill 
which is useful and well compensated outside of the scientific community 
(average salary of \$110k in the Washington, DC, area according to indeed.com at 
% RJH: Is this for a python developer?  Again, this is useless info, imho.
the time of writing). Python continues to see increased use in the astronomy 
community \citep{greenfield2011} which has similar goals and requirements as 
the solar physics community. Finally, Python integrates well with many 
technologies such as web servers \citep{dolgert2008}, SQL databases, 
interactive and shareable notebook-based computing \citep{perez2007}. For all 
of these reasons, Python is an ideal platform for a modern data analysis environment 
for solar physics.
